# Pipelining with muPipeR
*February 22, 2016*

##  Installation and set up
`muPipeR` can be directly installed from [github](https://github.com/MPIIComputationalEpigenetics/muPipeR)
```{r eval=FALSE}
library(devtools)
install_github("MPIIComputationalEpigenetics/muPipeR")
```

## Preliminaries
Loading the `muPipeR` package
```{r warning=FALSE, message=FALSE}
library(muPipeR)
# library(devtools)
# setwd("~/BroadSVN/eclipse_workspace/muPipeR")
# load_all("muPipeR")
```

Setting the analysis directory
```{r}
# anaDir <- "~/tmp_work/muPipeR_test"
anaDir <- file.path(tempdir(), "muPipeR_test")
```

## Pipelines

### Setting up the pipeline
Create the pipeline object
```{r message=FALSE}
pipe <- PipR(anaDir)
```

Note that the analysis directory must be non-existing. Creating a `PipR` object will create the analysis directory with its subdirectories:

directory | description
--- | ---
`results` | Contains the results of the analysis. For each analysis step, a subdirectory will be created that contains the results of that particular step.
`log` | Contains log files for the analysis jobs that are run.
`status` | Contains status information on the analysis.
`temp` | Contains temporary files.

Add an analysis step. Here, we simply create an empty input file
```{r}
pipe <- addStep(pipe, 
  "create_input_01",
  "touch",
  args="${STEPDIR}/input01.txt"
)
```

The following variables will be replaced in argument strings

variable | description
--- | ---
`${STEPDIR}` | result directory of the given analysis step
`${STEPDIR:stepname}` | result directory of a specific analysis step with name `stepname`

Add an analysis step with multiple jobs. Here, we create 5 additional empty files
```{r}
pipe <- addStep(pipe, 
  "create_input_multiple",
  rep("touch", 5),
  args=lapply(LETTERS[1:5], FUN=function(x){paste0("${STEPDIR}/input", x, ".txt")})
)
```

The next step takes an input file and appends a string to obtain its output file. This step depends to the input step to be successfully completed. This can be specified in the `parents` argument.
```{r}
# a shell script for appending a string to a file
appendTool <- system.file(file.path("extdata", "examples", "appendLine.sh"), package="muPipeR")
pipe <- addStep(pipe, 
  "append_01",
  appendTool,
  args=c(
    "${STEPDIR:create_input_01}/input01.txt",
    "${STEPDIR}/output01.txt",
    "'appended line 01'"
  ),
  parents="create_input_01"
)
# add a step appending a line to each of the input files
# create a list of arguments
aa <- lapply(1:5, FUN=function(i){
  c(
    paste0("${STEPDIR:create_input_multiple}/input", LETTERS[i], ".txt"),
    paste0("${STEPDIR}/output", LETTERS[i], ".txt"),
    paste0("'appended line ", LETTERS[i], "'")
  )
})
pipe <- addStep(pipe, 
  "append_multiple",
  appendTool,
  args=aa,
  parents="create_input_multiple"
)
```

Add an analysis step combining the output from multiple steps:
```{r}
# a shell script for concatenating multiple files (2nd to nth argument)
# into one output file (1st argument)
concatFilesTool <- system.file(file.path("extdata", "examples", "concatFiles.sh"), package="muPipeR")
pipe <- addStep(pipe, 
  "join_01_ABC",
  concatFilesTool,
  args=c(
    "${STEPDIR}/concat_01ABC.txt",
    "${STEPDIR:append_01}/output01.txt",
    "${STEPDIR:append_multiple}/outputA.txt",
    "${STEPDIR:append_multiple}/outputB.txt",
    "${STEPDIR:append_multiple}/outputC.txt"
  ),
  parents=c("append_01", "append_multiple")
)
```

### Inspecting the pipeline
Show the pipeline graph
```{r}
plotGraph(pipe)
```

### Running the pipeline
All steps in the pipeline are executed using the `run` command:
```{r}
pipe <- run(pipe)
```

In the output object, the steps are marked as completed (e.g. they are green in the graph):
```{r}
plotGraph(pipe)
```


### Resetting pipeline steps
Undo/reset a pipeline step and its dependent steps:
```{r}
pipe <- resetStep(pipe, "append_01")
plotGraph(pipe)
```

Per default this will also delete the corresponding subdirectories in the result directory.
If you run the pipeline again, the reset steps will be run again:
```{r}
pipe <- run(pipe)
```

### Cleaning up
To remove temporary files and empty log files use:
```{r}
cleanup(pipe)
```

## Running jobs using `CommandR` and its subclasses
The `CommandR` class is a virtual class implementing methods for running commands along with corresponding arguments.
Currently there are two child classes implemented in the package:

class | description
--- | ---
`CommandRsystem` | A class for running commands using command line calls. More specifically, the `system2` function implemented in `R` is run to send a given combination of cammand and arguments to the command line.
`CommandRsge` | A class for submitting commands to a compute cluster running a Sun Grid Engine (SGE). Jobs are submitted using calls to `qsub`. So, the executing script must be run on a machine capable of submitting jobs to the cluster.

```{R}
cmdr <- CommandRsystem()
```

`CommandR` objects can also be initialized with a logging directory where the output of the commands is written to:
```{R}
cmdr <- CommandRsystem(logDir=tempdir())
```

Note, that for `CommandRsge` objects, it is mandatory to specify a logging directory that can be accessed by all execution nodes in the SGE queue.

Get the logging directory:
```{R}
getLogDir(cmdr)
```

In order to run a command with a `CommandR`, it has to be packaged into a `Job` object. 
```{R}
echoJob <- Job("echo", args="'Hello!'")
```

If you want to identify a job later, it might also be useful to name it:
```{R}
echoJob <- Job("echo", args="'Hello!'", id="say_hello")
```

You can access the command that will be executed, it's arguments and other job details:
```{R}
getId(echoJob)
getCmd(echoJob)
getArgs(echoJob)
getCallString(echoJob)
```

The `exec` method can be used to actually execute jobs:
```{R}
jobRes <- exec(cmdr, echoJob)
```

If you want the output of a job to be stored in the result, use the `result` argument:
```{R}
jobRes <- exec(cmdr, echoJob, result=TRUE)
```

The result of the execution is an object of class `JobResult`, that contains information on the job's output, errors and status:
```{R}
getOut(jobRes)
getStatus(jobRes)
getErr(jobRes)
```

You can also execute an array of jobs using `lexec` (list execute):
```{R}
echoJobList <- lapply(1:5, FUN=function(i){
  Job("echo", args=paste0("'Hello ", i, "!'"), id=paste0("say_hello_", i))
})
jobResList <- lexec(cmdr, echoJobList)
```

If you want to specify that a job depends on other jobs you can add them using the `dependsOn` argument:
```{R}
echoJob2 <- Job("echo", args="'I mean: Hello, my name is HAL2000.'", 
                id="say_hello_2", dependsOn=list(echoJob))
```

---
author: *Fabian Müller* <<fmueller@mpi-inf.mpg.de>>
